---
title: "Extended Figure Data 10"
author: "dsarni"
date: "04-03-2026"
output: html_document
---

## Extended Data Figure 10. Examples of hypermethylated transcription factor promoter regions; accelerated epigenetic clock ageing in HESJAS

1. Libraries used in this figure.

```{r}
library(ggplot2)
```

2. Import data.

```{r}
# EDF10.e
human_epiClocks <- read.table("../data/EDF10/human_epiclock_hesjas.tsv", header = T)

# EDF10.f
mouse_epiClock <- read.csv("../data/EDF10/mouse_epiclock.csv")
```

### EDF10.e

3. Plot epiclock age predictions for HESJAS patients and controls

```{r}
ggplot(human_epiClocks, aes(x = variable, y = value, color = Geno)) +
  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.4) + #width = 0.2,
  stat_summary(aes(group = Geno), fun = median, geom = "crossbar",
               position = position_dodge(width = 0.6),
               width = 0.5, fatten = 2, color = "grey20") +
  geom_vline(xintercept = seq(1.5, length(unique(human_epiClocks$variable)) - 0.5, 1),
             linetype = "dashed", color = "grey75") +
  geom_hline(yintercept = 0, color = "grey65") +
  theme_classic(base_size = 14) +
  theme(legend.title = element_blank()) +
  labs(x = "Epiclock", y = "Delta age pred.-chron. (years)") +
  scale_color_manual(values = c("Control" = "royalblue3", "Patient" = "tomato3"))
```

4. Compute *p-values*

* Hannum clock

```{r}
# Hannum
hannum_df <- subset(human_epiClocks, variable == "Hannum")

t.test(value ~ Geno, data = hannum_df, var.equal = F)
```

* Horvath clock

```{r}
# Horvath
horvath_df <- subset(human_epiClocks, variable == "Horvath")

t.test(value ~ Geno, data = horvath_df, var.equal = F)
```

* PhenoAge clock

```{r}
# PhenoAge
phenoAge_df <- subset(human_epiClocks, variable == "PhenoAge")

t.test(value ~ Geno, data = phenoAge_df, var.equal = F)
```

### EDF10.f

5. Convert the age delta predictions from days to months

```{r}
mouse_epiClock$delta_month <- mouse_epiClock$Age_delta/30.5
```

6. Plot mouse epiclock age predictions delta

```{r}
ggplot(data = mouse_epiClock, aes(x = Condition, y = delta_month, color = as.factor(Condition)))+
  geom_jitter(size = 4, alpha = 0.5, width = .2)+
  scale_color_manual(values = c("Control" = "royalblue3", "Mutant" = "tomato3"))+
  ylab("D age pred.-chron. (month)")+
  xlab("")+
  ylim(c(0,20))+
  stat_summary(fun = median, geom = "crossbar", color = "black", width = 0.5, linewidth = 0.5, fatten = 2, show.legend = T)+
  theme_classic()+
  theme(text = element_text(size = 14),
        axis.text = element_text(size = 14),
        legend.position = "none")
```

7. Compute *p-values*

```{r}
t.test(delta_month ~ Condition, data = mouse_epiClock, var.equal = F)
```

```{r}
sessionInfo()
```