---
title: "Sarni et al., Figure 6"
author: "dsarni"
date: "26-02-2026"
output: html_document
---

## Figure 6. DNMT3AGOF hypermethylation corresponds to sites of DNA methylation gains in human ageing

1. Libraries used in this figure.

```{r, message = F}
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

2. Import files

```{r}
# Figure 6.b
heyn_k27_high <- read.table("../data/Figure_6/Heyn_PNAS_hDMVs_k27_HIGH_hg19_20K_10k_250_211124.tsv", header = T)
heyn_k27_low <- read.table("../data/Figure_6/Heyn_PNAS_hDMVs_k27_LOW_hg19_20K_10k_250_211124.tsv", header = T)

# Figure 6.c
gen_scot_df <- read.table("../data/Figure_6/Gen_Scot_corr_hesjas_horvath.tsv.gz", header = T)
```

### Fig6.b

DNA methylation profiles along hDMVs in three healthy individuals from [Heyn et al., PNAS 2013] (https://doi.org/10.1073/pnas.1120658109).

3. Split tables by samples (age)

```{r}
# K27.high
nb.k27.high <- heyn_k27_high[,1:160]
mid.k27.high <- heyn_k27_high[,161:320]
old.k27.high <- heyn_k27_high[,321:480]

# K27.low
nb.k27.low <- heyn_k27_low[,1:160]
mid.k27.low <- heyn_k27_low[,161:320]
old.k27.low <- heyn_k27_low[,321:480]
```

4. Create pileup tables to fit plotting with ggplot

```{r}
##################
# hDMV at high k27
##################

pileup_heyn_df_high <- data.frame(nb = NA,
                                  mid = NA,
                                  old = NA)

for(i in 1:length(nb.k27.high)){
  
  pileup_heyn_df_high[i,1] <- mean(nb.k27.high[,i], na.rm = T)
  pileup_heyn_df_high[i,2] <- mean(mid.k27.high[,i], na.rm = T)
  pileup_heyn_df_high[i,3] <- mean(old.k27.high[,i], na.rm = T)
  
}

pileup_heyn_df_high$X <- as.character(row.names(pileup_heyn_df_high), ordered = T)
pileup_heyn_df_high$X <- factor(pileup_heyn_df_high$X, levels = unique(pileup_heyn_df_high$X))

#################
# hDMV at low k27
#################

pileup_heyn_df_low <- data.frame(nb = NA,
                                 mid = NA,
                                 old = NA)

for(i in 1:length(nb.k27.low)){
  
  pileup_heyn_df_low[i,1] <- mean(nb.k27.low[,i], na.rm = T)
  pileup_heyn_df_low[i,2] <- mean(mid.k27.low[,i], na.rm = T)
  pileup_heyn_df_low[i,3] <- mean(old.k27.low[,i], na.rm = T)
  
}

pileup_heyn_df_low$X <- as.character(row.names(pileup_heyn_df_low), ordered = T)
pileup_heyn_df_low$X <- factor(pileup_heyn_df_low$X, levels = unique(pileup_heyn_df_low$X))
```

5. Plots

* Polycomb positive hDMVs

```{r}
ggplot(data = pileup_heyn_df_high, aes(x=X))+
  
  geom_line(aes(y=nb, colour = "nb", group =1), linewidth = 1)+
  geom_line(aes(y=mid, colour = "mid", group =1), linewidth = 1)+
  geom_line(aes(y=old, colour = "old", group =1), linewidth = 1)+
  
  scale_x_discrete(breaks = c(1, 41, 121, 160), labels = c("-10kb", "start", "end", "+10kb")) +
  scale_colour_manual(name = "DMVs", 
                      values = c("#ff9999","#ff3333","#990000"),
                      breaks=c("nb","mid","old"),
                      labels=c("nb","mid","old"))+
  xlab("")+
  ylab("DNAme (M/T ratio)")+
  ylim(c(0,0.7))+
  theme_bw()+
  theme(text = element_text(size = 16),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
dev.off()
```

* Polycomb positive hDMVs

```{r}
ggplot(data = pileup_heyn_df_low, aes(x=X))+
  
  geom_line(aes(y=nb, colour = "nb", group =1), linewidth = 1)+
  geom_line(aes(y=mid, colour = "mid", group =1), linewidth = 1)+
  geom_line(aes(y=old, colour = "old", group =1), linewidth = 1)+
  
  scale_x_discrete(breaks = c(1, 41, 121, 160), labels = c("-10kb", "start", "end", "+10kb")) +
  scale_colour_manual(name = "DMVs", 
                      values = c("#66ffb2","#2AD82A","#00994C"),
                      breaks=c("nb","mid","old"),
                      labels=c("nb","mid","old"))+
  xlab("")+
  ylab("DNAme (M/T ratio)")+
  ylim(c(0,0.9))+
  theme_bw()+
  theme(text = element_text(size = 16),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

### Fig6.c

Generation Scotland DNA methylation values correlation with age

6. Plot violin plots for CpG correltation with age. Use HESJAS gain DMR associated CpGs and Horvath's epiclock CpGs.

```{r}
ggplot(data = gen_scot_df, aes(x = type, y = cor, fill = type))+
  geom_violin(alpha = 0.75)+
  scale_fill_manual(values = c("#A8A8A8","#FF4343","#FF8000"))+
  geom_boxplot(width = 0.1, fill = "white", outliers = F)+
  geom_hline(yintercept = 0, lty = 3)+
  scale_x_discrete(limits=c("control","horvath","dmr"))+
  theme_classic() +
  labs(x = "", y = "Pearson correlation")+
  theme(legend.title = element_blank())
```

7. Calculate the p-value

```{r}
pairwise.wilcox.test(gen_scot_df$cor, gen_scot_df$type,
                     p.adjust.method = "BH")
```


```{r}
sessionInfo()
```