---
title: "Sarni et al., Figure 1"
author: "dsarni"
date: "22-02-2026"
output: html_document
---

## Figure 1. HESJAS syndrome

1. Library used for this figure.

```{r}
library(RColorBrewer)
```

2. Import data - DNA methylation in control and HESJAS patients within DNA hypermethylated DMRs.

Input data for this figure will be processed and normalised beta values for CpGs within HESJAS hypermethylated DMRs.

```{r}
# Import data table

hDMR_blood_beta <- read.csv("../data/Figure_1/Figure_1ef_hesjas_gain_dmr_meth.csv")
```

3. Compute the mean for control samples to be used for ordering the rows in the heatmap.

```{r}
hDMR_blood_beta.mean.ctrl <- apply(hDMR_blood_beta[,c(2:10)], 1, mean, na.rm=TRUE)
```

### Figure 1.e

```{r}
image(x=c(1:ncol(hDMR_blood_beta[,2:17])),y=c(1:nrow(hDMR_blood_beta)),
      z=t(hDMR_blood_beta[order(hDMR_blood_beta.mean.ctrl), 2:17]),
      col=rev(brewer.pal(10,"RdYlBu")),breaks=seq(0,1,by=0.1),
      axes=FALSE,xlab="",ylab="DMRs with DNAme gain (n=1,183)")
box()
mtext(colnames(hDMR_blood_beta[,2:17]), line=0.5, side=1, at=c(1:16),
      cex=0.6, las=2)
```

### Figure 1.f

```{r}
# For boxplots plot the mean beta values (DNA methylation) per condition, columns 18:19 in the data table

boxplot(hDMR_blood_beta[,18:19],
        names=rep("", times=2), range=0, ylim=c(0,1), col = c(("#4169C4"),("#FF6666")),
        ylab="Beta Value", cex.main=1, main="EPIC - blood samples in up blood up.DMRs")
mtext(colnames(hDMR_blood_beta)[18:19], line=0.5, side=1, at=c(1:2),
      cex=0.6, las=2)
```

4. Compute the *P-value* for Figure 1.f using Wilcox.test.

```{r}
wilcox.test(hDMR_blood_beta[,18],hDMR_blood_beta[,19])
```

```{r}
sessionInfo()
```