---
title: "Sarni et al., Figure 2"
author: "dsarni"
date: "24-02-2026"
output: rmarkdown::github_document
---

## Figure 2. Dnmt3a w326r/+ mice model HESJAS syndrome and exhibit multiple age-associated phenotypes

1. Libraries used in this figure.

```{r, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
```

2. Import data - DNA hypermethylated DMRs in bone marrow from +/+ and w326r/+ ages: 4-day, 23-day, 3-month and 1-year old mice.

```{r}
# Import data tables

bm <- read.csv("../data/Figure_2/Figure_2h_bm_all_dmr_mean_age.csv")
```

3. Convert the mean age data into a format suitable for ggplot

```{r}
bm.df <- data.frame(Time = rep(c(rep("4d", length(bm[[1]])),
                             rep("23d", length(bm[[1]])),
                             rep("3m", length(bm[[1]])),
                             rep("1yr", length(bm[[1]]))),2),
                    Geno = c(rep("+/+",length(bm[[1]])*4),
                             rep("W326R/+",length(bm[[1]])*4)),
                    Value = c(bm$c.4d,
                              bm$c.23d,
                              bm$c.3m,
                              bm$c.1yr,
                              bm$m.4d,
                              bm$m.23d,
                              bm$m.3m,
                              bm$m.1yr))

bm.df$Group <- interaction(bm.df$Geno, bm.df$Time, sep = "_")
```

4. Order samples for plotting.

```{r}
group_levels <- levels(factor(bm.df$Group))
group_levels <- c(group_levels[7],
                  group_levels[3],
                  group_levels[5],
                  group_levels[1],
                  group_levels[8],
                  group_levels[4],
                  group_levels[6],
                  group_levels[2])
group_levels
```

```{r}
bm.df$Group <- factor(bm.df$Group, levels = group_levels)
```

### Figure 2.h. Boxplot for mean DNA methylation per age per genotype

```{r}
ggplot(bm.df, aes(x = Geno, y = Value, fill = Group)) +
  stat_boxplot(
    position = position_dodge(0.8),
    geom = "errorbar",
    width = 0.4
  ) +
  geom_boxplot(position = position_dodge(width = 0.8), 
               colour = NA) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.75,
    colour = "black",
    position = position_dodge(0.8)
  ) +
  scale_fill_manual(values = c(
    "+/+_4d" = "#d4def1",
    "+/+_23d" = "#90bde5",
    "+/+_3m" = "#516fb5",
    "+/+_1yr" = "#292c6a",
    "W326R/+_4d" = "#f7b0b2",
    "W326R/+_23d" = "#f17275",
    "W326R/+_3m" = "#ec2f2d",
    "W326R/+_1yr" = "#9c1d22"
  )) +
  labs(x = "Genotype", y = "B-Value", fill = "Group") +
  theme_classic()
```

5. Compute the *P-values* between ages per genotype

* wild-type mice 

```{r}
# Kruskal-Wallis for +/+
bm.df.wt <- bm.df[bm.df$Geno == "+/+",]
  
kruskal.test(Value~Time, data = bm.df.wt)

pairwise.wilcox.test(bm.df.wt$Value, bm.df.wt$Time, p.adjust.methods = "BH")
```

* W326R/+ mice 

```{r}
# Kruskal-Wallis for W326R/+
bm.df.mut <- bm.df[bm.df$Geno == "W326R/+",]
  
kruskal.test(Value~Time, data = bm.df.mut)

pairwise.wilcox.test(bm.df.mut$Value, bm.df.mut$Time, p.adjust.methods = "BH")
```

Compute the *P-values* between genotypes (day 4)

```{r}
# Two-sided, paired Wilcoxon rank sum test for +/+ versus W326R/+
day4 <- subset(bm.df, Time == "4d")
pairwise.wilcox.test(day4$Value, day4$Geno, p.adjust.methods = "BH")
```

```{r}
sessionInfo()
```